Utilizing Google's Mobility Report to understand Infectious Disease Dynamics

### 2025-03-14 code/processing-code/processingfile-v1.qmd
Part 3 of the project
Done modeling and some work on manuscript, Need to define workflow for analysis and finish on supplementary analyses.

```{r}
# load needed packages
library(here) 
library(dplyr)
library(skimr)
library(ggplot2)
library(zoo)
library(tidyr)
```

@@

@@

@@

>>>>

```{r}
# load needed packages
library(here) 
library(dplyr)
library(skimr)
library(ggplot2)
library(zoo)
library(tidyr)
library(lubridate)
library(readxl)
```

@@

# county population information
ch_pop_data <- read_excel(here::here("data", "raw-data", "co-est2023-pop.xlsx"), skip = 3)
```

```{r}
# Remove the second column
ch_pop_data <- ch_pop_data[, -2]

ch_pop_data <- ch_pop_data %>%
  pivot_longer(
    cols = c(`2020`,`2021`, `2022`, `2023`),    # Specify the year columns
    names_to = "year",                    # Create a new column for the years
    values_to = "population_count"              # Create a new column for population counts
  )

# Fix the first column
ch_pop_data <- ch_pop_data %>%
  mutate(
    county = str_remove(`...1`, "^\\."),            
    county = str_remove(county, "County"),     
    county = str_to_lower(county),                  
    state = str_extract(county, "(?<=,).*"),        
    county = str_remove(county, ",.*")              
  ) %>%
  mutate(
    state = str_to_lower(str_trim(state))           
  )

# Change year column to be double for joining later
ch_pop_data$year <- as.double(ch_pop_data$year)

# Clean names to prepare for joining
ch_pop_data <- ch_pop_data %>%
  mutate(
    county = str_trim(str_to_lower(county)),  
    state = str_trim(str_to_lower(state))     
  )

@@

This section looks through the CHR data

```{r}
# First need to filter out the data for what years we are interested in
unique(chr_data_2024$yearspan)

# Based on this, we are most interested in 2020-2022 because of its overlap with the other data.
chr_data_filtered <- filter(chr_data_2024, yearspan %in% c("2020", "2021", "2022"))

# Change yearspan to be a double. This is important for joining all the data frames later.
chr_data_filtered$yearspan <- as.double(chr_data_filtered$yearspan)

# Now we need to decide what meausre interest us the most
unique(chr_data_filtered$measurename)

# We chose uninsured adults as an indicator of access to healthcare (this is listed in the website)
chr_data_filtered <- filter(chr_data_filtered, measurename == "Uninsured adults")

# rename the counties to prep for dataframe joining
chr_data_filtered <- chr_data_filtered %>%
  mutate(
    county = tolower(county),  # Convert 'county' to lowercase
    county = gsub(" county", "", county), # removes the word county
    uninsured_rate = (numerator / denominator) * 100000) # uninsured rate per 100,000

# Check if counties have same names across different staes
duplicates <- chr_data_filtered %>%
  group_by(county, state) %>%
  tally() %>%
  filter(n > 1)

# View the duplicates and looks like there is... need to deal with this if I want to join the data with the other data
print(duplicates)

# Change abbreviated state names into the full name so I can join them easier

# Create a mapping for state abbreviations to full names and convert the full names to lowercase
state_mapping <- data.frame(
  state_abbr = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY"),
  state_full = tolower(c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware", "Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming"))
)

# Convert state abbreviation to full name in chr_data_2024
chr_data_filtered <- chr_data_filtered %>%
  left_join(state_mapping, by = c("state" = "state_abbr"))

@@

merged_df_clean <- merged_df_clean %>%
  mutate(year = year(date)) %>%
  left_join(chr_data_filtered %>% select(state_full, county, uninsured_rate, yearspan), 
            by = c("state" = "state_full", "county" = "county", "year" = "yearspan"))

merged_df_clean <- merged_df_clean %>%
  left_join(
    ch_pop_data %>% 
      select(year, population_count, county, state),
    by = c("state" = "state", "county" = "county", "year" = "year")
  )
```

### 2025-03-14 products/manuscript/Manuscript.qmd
Part 3 of the project
Done modeling and some work on manuscript, Need to define workflow for analysis and finish on supplementary analyses.

---
title: "Utilizing Google's Mobility Report to understand Infectious Disease Dynamics"
author: "Vincent Nguyen"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

Some text was left from the template in order to adjust to any feedback given. The manuscript will be more fleshed out once the final project idea is approved.

@@

-   Vincent Nguyen$^{1}$ (ORCID: 0000-0000-1234-5678)

@@

## General Background Information

Understanding the spread of infectious diseases is crucial for public health preparedness and response. Disease infection dynamics are shaped by multiple factors, including population movement, the pathogen itself, and public health interventions. Mobility is considered a key driver of human interactions and thus a primary vehicle for transmission potential. The COVID-19 pandemic provides a great example of how changes in mobility, driven by policies and public health measures, influenced infection rates.

@@

The CDC Vaccination hesitancy survey is a possible data set I could use to make the analysis more complex.\
\
\

>>>>

---
title: "Utilizing Google's Mobility Report to understand Infectious Disease Dynamics"
author: "Vincent Nguyen"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/MADA.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

@@

-   Vincent Nguyen$^{1}$

@@

The emergence of the novel corona-virus SARS-CoV-2 caused one of the most significant global health crises in modern history. First identified in Wuhan, China, in late 2019, the virus’s transmissibility and globalism led to its rapid spread worldwide, to leading the World Health Organization to declare it a pandemic on March 11, 2020. Just two days later, on March 13th, 2020, the Trump Administration declared a nationwide emergency due to the COVID-19 pandemic, marking the biggest virus outbreak since \_\_\_. In response, public health agencies implemented measures to curb the virus’ spread including travel restrictions, social distancing, and lockdown procedures (CDC 2024)[@centersfordiseasecontrolandprevention_2024_covid19].

\[insert stats about pandemic\]

### 2.2 Lockdown and Social Distancing

Nonpharmacological interventions are among the measures public health officials utilize to help curb the transmission of infectious diseases. In lieu of medication or vaccinations, these interventions can be a cost-effective and population-based solution. These interventions include the use of isolation, the subdivision of an infected individual from the general population, and social distancing, a form of voluntary or mandated physical distancing from others. Another common measure is the use of quarantine, which is the restriction of infected or at-risk individuals from traveling to help minimize the risk of healthy individuals; it is intended that the period of transmissibility has passed before lifting the quarantine (Jacobsen and Jacobsen 2020)[@jacobsen_2020_statewide].

Uniquely, the COVID-19 pandemic utilized a new form of physical distancing measures which was known as stay-at-home orders and colloquially, lockdowns. Lockdowns involved stringent stay-at-home orders, closure of non-essential businesses, and restrictions on public gatherings. By April 3rd, 2020, just a few weeks after California’s state-wide mandate, 41 states joined the directive in restricting mobility (Mervosh et al. 2020)[@mervosh_2020_see]. Many polictians and citizens questioned the measure citing a lack of evidence for its efficacy; looking at the Wikipedia page shows a lack of prior history implementing lockdowns (Zeleny et al. 2021)[@zeleny_2021_a].

### 2.3 Mobility Dynamics

According to Paltra et al. 2021, mobility reductions on COVID-19 can be quadratic in reducing the transmission. For example, the authors state a reduction of out-of-home duration to 70% can result in a transmission rate of 0.49. \[finish\]

@@
